---
title: "Arrays with Fixed Margins in Stan"
author: "Gidon Cohen"
format: html
bibliography: references.bib
editor: source
execute: 
  echo: false
  message: false
  warning: false
---

```{r}
#| echo: false
#| warning: false
#| message: false


library(rstan)
library(tidyverse)
library(janitor)
library(huxtable)


```

In this note I describe some approaches to modeling matricies and arrays with fixed row and column margins in Stan where cell counts must be positive. To assist with modelling I relax the assumption that cell counts must be integers. Integer cell counts summing exactly to margin constraints are obtained in generated quantities using integer versions sequential sampling algorithms described below. The approaches works by making allocations sequentially, e.g. to each cell in turn, where each allocation ensures that the realisation of an overall table or array summing to the fixed margin remains possible. 

To illustrate some of the advantages of sequential sampling I also describe a raking approach to realising tables with exact margins. 

1.  Exact margin approach. This approach is like a two-way stick breaking algorithm. It uses ideas from Sequential Importance Sampling (Chen et. al 2005) to build an algorithm based on an unconstrained allocation space to produce cell counts that ensure that both margin constraints are met. Thus, $(R - 1)(C - 1)$ unconstrained allocation parameters ($\lambda$) are transformed to RxC cell values which sum precisely to row and column margins. Because the transform is sequential, the Jacobian is quite straightforward.  
The resulting cell counts can be modeled in various ways including for example by further transforming the cell counts to the simplex underlying the table. In the examples below I further transform the cell values to an RxC - 1 simplex representation of the table ($\text{ilr}$) in order to provide structured information about the type of table to be generated. 

## Basis of the sequential sampling approaches

Fréchet bounds describe the maximum and minimum value of each cell in a two-way table with total $T$, row margins $R_1 \dots R_R$ and column margins $C_1 \dots C_C$:

$$ \mathrm{max}\left(0, R_r + C_c - T \right) \le \pi_{rc} \le \mathrm{min} \left( R_r, C_c \right)$$

Given these bounds, @Chen2005 describe a procedure for sequentially allocating values to cells that constructs only valid two-way contingency tables with fixed margins. Their approach begins the allocation from the top-left cell in the table ($\pi_{11}$). With no cells allocated, the limits on $\pi_{11}$ are:

$$ \mathrm{max}\left(0, R_1 + C_1 - T \right) \le \pi_{11} \le \mathrm{min} \left( R_1, C_1 \right)$$

This is the only condition that $\pi_{11}$ needs to satisfy. Recursively, suppose that we have chosen $\pi_{i1}= \pi^{\prime}_{i1}$ for $1 \le i \le r-1$. Then the only restriction on $\pi_{r1}$ is
$$ \mathrm{max}\left( 0, \left(C_1 - \sum_{i=1}^{r-1} \pi_{i1}^{\prime} \right) - \sum_{i=r+1}^{R} R_i\right) \le \pi_{r1} \le \mathrm{min}\left(R_r, C_1 - \sum_{i=1}^{r-1} \pi_{i1}^{\prime}\right)$$

This describes a process for sequentially calculating cell bounds from the top of the first column down to the bottom of that column. If the first column of a contingency table is fully allocated (and only those cells are allocated) then the remaining unallocated cells in columns $2 \dots C$ form a contingency table of dimensions $R \times (C-1)$ where the row margins are $R_1 - \pi_{11}, \dots, R_R - \pi_{R1}$. Hence, we can continue the allocation from the top cell of each column once the previous column is fully allocated.

In general, when the cell $\pi_{rc}$ is reached, let $R^{\prime}_i = R_i - \sum_{j=1}^{c-1} \pi^{\prime}_{ij}$ be what is left of row $i$ and $C^{\prime}_{c} = C_c - \sum_{i=1}^{r-1} \pi^{\prime}_{ic}$ what is left of column $c$. The lower bound $\alpha_{rc}$ and upper bound $\beta_{rc}$ of the cell are then
$$\alpha_{rc} = \mathrm{max}\left( 0, C^{\prime}_{c} - \sum_{i=r+1}^{R} R^{\prime}_i\right), \qquad \beta_{rc} = \mathrm{min}\left(R^{\prime}_r, C^{\prime}_{c}\right),$$
and can be calculated sequentially. (The Stan code fills the table row by row rather than column by column, which is the same construction with rows and columns swapped.)

### Dealing with Reduced Degrees of Freedom Arising from Zeros in the Row or Column Margins

In a case of an RxC contingency table with fixed margins where none of the margins are equal to zero, then there are $(R - 1)(C - 1)$ degrees of freedom, which defines the number of parameters needed to fully allocate the table by the sequential sampling. In the case that some of the margins are equal to zero, then this reduces the degrees of freedom because all of the cells in any row of column with a margin sum of zero are required to be zero. To deal with the reduced number of parameters required, I make the allocations in a reduced dimension matrix (with zero rows and columns removed), and then increase to the RxC dimensions by inserting rows and/or columns of zero counts into the cells which make up the rows and/or columns that sum to zero.


## Representations of the table simplex

One finding from these models is that even when descriptions of simplexes have wide equivalence the way in which they are parameterised can make a big difference to Stan.

## The four allocation functions

Three of the four allocation functions build on the sequential transformation in @Chen2005, in that they provide different ways of choosing a point for a cell between its lower and upper feasible bounds. The fourth, adjusted table, instead solves for an adjustment to the whole table.

### Position logit (between bounds)

In the between bounds transformation each of the $(R - 1)(C - 1)$ unconstrained allocation parameters $\lambda_{rc}$ places a cell within its feasible range: $\text{logit}^{-1}(\lambda_{rc})$ is the cell's position between the lower and upper bounds, as a share of the width of that range. The transformation operates sequentially starting from $\pi_{11}$. Each parameter $\lambda_{rc}$ gives a cell value $\pi_{rc}$ between the lower bound ($\alpha_{rc}$) and upper bound ($\beta_{rc}$) possible for the cell given the row and column margins and any allocations already made to other cells.

For rows $1 \dots R-1$ and columns $1 \dots C-1$ the unconstrained parameter $\lambda_{rc}$ is transformed to a value $\pi_{rc}$ in the range from $\alpha_{rc}$ to $\beta_{rc}$ with the inverse logit function: $$\pi_{rc} = \alpha_{rc} + \frac{1}{1+e^{-\lambda_{rc}}}(\beta_{rc} - \alpha_{rc})$$

For cells in column $C$ of rows $1 \dots R-1$ the value is given by the remaining row total left to be allocated. For cells in row $R$ the value is given by the remaining column total left to be allocated.

This transformation gives cell values which are real numbers greater than or equal to 0. This continuous version is needed for the `stan` modelling, and can be thought of as generating something like an expected value of the cell. The algorithm can be amended to give cells as integer values, by rounding the transformed value to the nearest integer:

$$\pi_{rc} = \left\lfloor \alpha_{rc} + \frac{1}{1+e^{-\lambda_{rc}}}(\beta_{rc} - \alpha_{rc}) \right\rceil$$

Stan requires functions to be differentiable, so a continuous function can be used to approximate the lower bound $\alpha_{rc}$ and the upper bound $\beta_{rc}$. In earlier versions I adapted the continuous hinge functions described by Andrew Gelman on his [blog](https://statmodeling.stat.columbia.edu/2017/05/19/continuous-hinge-function-bayesian-modeling/).

<!-- CHECK: the smoothing in stan/functions/alloc_bounds.stan (delta > 0) is not the Gelman hinge below; it rounds each kink inward over about delta x the cell's range. Decide which to describe. -->

For the lower bound I approximate $\mathrm{max}\left(0, C_1 + R_1 - T \right)$ with the function
$$\alpha = \delta \log \left(1 +  e^{\frac{C_1 + R_1 - T}{\delta}} \right)$$
As $C_1 + R_1 - T$ falls below zero this function quickly approaches zero; when it is greater than zero the function approximates $C_1 + R_1 - T$. $\delta$ is a parameter which determines how quickly that happens (smaller values give a sharper approximation to the max function).

Similarly, I approximate the upper bound $\mathrm{min} \left( R_1, C_1 \right)$ with the function
$$ -\delta \log \left(e^{\frac{-R_1}{\delta}} + e^{\frac{-C_1}{\delta}} \right)$$

Again $\delta$ determines how quickly that happens (smaller values give a sharper approximation to the min function).

#### The log determinant of the Jacobian of the transform for between bounds

Many of the Jacobian components are zero, and the sequential nature of the transform means that the Jacobian matrix is triangular. The elements on the diagonal simplify to a convenient function of bound widths and allocation parameters. If we re-index the $L = (R - 1)(C - 1)$ parameters, the upper bounds ($\beta$), the lower bounds ($\alpha$) and the allocation parameters ($\lambda$) using $l = r + (c - 1)(R - 1)$, then:

$$\ln (\det J_{f-1}(y)) = \ln \left( \left(\beta_1 -\alpha_1 \right) \cdot \textrm{logit}^{-1}(\lambda_1) \cdot (1 - \textrm{logit}^{-1}(\lambda_1))\right) + 
\dots + 
\ln \left( \left(\beta_L -\alpha_L \right) \cdot \textrm{logit}^{-1}(\lambda_L) \cdot (1 - \textrm{logit}^{-1}(\lambda_L))\right)$$

### Odds-ratio logit (cell logit)

The odds-ratio logit uses the same sequential fill as the position logit, but each allocation parameter is the log odds ratio of a $2 \times 2$ table, not a position between bounds. When cell $\pi_{rc} = x$ is reached, collapse the rest of the table around it: let $s_R$ be what is left of the row, $s_C$ what is left of the column, and $\rho$ what is left of the columns the row has not yet visited. The cell sits in the table

$$\begin{pmatrix} x & s_R - x \\ s_C - x & \rho - s_R + x \end{pmatrix}$$

and its log odds ratio is
$$\lambda = \log x + \log(\rho - s_R + x) - \log(s_R - x) - \log(s_C - x).$$

The odds ratio increases with $x$, so for any $\lambda$ there is exactly one $x$ strictly inside the feasible range, found as the root of a quadratic. As $\lambda$ runs over the real line $x$ covers the whole range from $\alpha_{rc}$ to $\beta_{rc}$, and $\lambda = 0$ is independence: $x = s_R s_C / (s_C + \rho)$. As with the position logit the Jacobian is triangular, with the derivative of $x$ with respect to $\lambda$ on the diagonal. The difference is in the geometry: the position logit's midpoint is the middle of the feasible range and its bounds have kinks (where the binding bound changes), whereas here the cell is a smooth function of $s_R$, $s_C$ and $\rho$, so no smoothing parameter $\delta$ is needed.

### Adjusted row

The adjusted row function allocates a whole row in one step instead of one cell at a time. For the row being filled, let $\kappa_c$ be the capacity left in column $c$ after earlier rows. The cells of the row are

$$x_c = \kappa_c \cdot \text{logit}^{-1}(\eta_c + \tau),$$

where the $\eta_c$ are free parameters (one column, the reference column, has $\eta_c = 0$, which leaves $C - 1$ free parameters per row) and $\tau$ is a single adjustment that is solved so that the row adds up to its total. The row total is increasing in $\tau$, so this is a one-dimensional monotone solve. Rows are filled in turn and the last row is found by subtraction, giving $(R-1)(C-1)$ free parameters in all.

Each cell lies strictly between zero and its column's remaining capacity, the row total is met exactly, the whole feasible range can be reached and there are no kinks. The solve for $\tau$ is implicit, but the Jacobian of a row with respect to its $\eta$ parameters has a closed form (a rank-one adjustment of a diagonal matrix), and the Jacobian of the whole table is block triangular across rows.

### Adjusted table

The adjusted table function allocates the whole table at once, in the manner of raking (iterative proportional fitting). The cells are

$$\pi_{rc} = \exp(\theta_{rc} + u_r + v_c),$$

where $\theta$ holds the $(R-1)(C-1)$ free parameters, which are log odds ratios against the last row and last column ($\theta$ is zero in the last row and column), and $u_r$ and $v_c$ are one adjustment per row and per column, solved jointly so that both sets of margins are met. The table is exactly the one that raking converges to when started from the seed $\exp(\theta)$.

The adjustments minimise a convex function, so for any $\theta$ they exist and are unique (up to the one redundant direction between $u$ and $v$, fixed by setting $v_C = 0$). There is no fill order and there are no kinks. The price is that each evaluation requires solving a system of $R + C - 1$ equations (a few raking sweeps and then Newton steps), and the Jacobian is found by implicit differentiation, so it is not triangular and its cost grows more quickly with table size than for the sequential functions.

## The four allocation functions work 
In this section I show that the there is a one-to-one map from RxC onto the tables with specific margins.

Each allocation function should be a one-to-one map from an unconstrained
vector of $K = (R-1)(C-1)$ parameters onto the interior of the set of
$R \times C$ tables with the given margins, with a Jacobian we can write
down. We checked this for each scheme at table sizes from $2 \times 2$ to
$7 \times 7$, at random margins and parameters, and for two fill orders (rows
in order with the last column by subtraction, and a random row order with a
random subtraction column in each row). The checks were:

- **margins**: the largest relative error in a row or column total;
- **smallest cell**: the smallest cell, as a share of the table total
  (it should be positive);
- **inverse**: the largest error in recovering the parameters from the table;
- **Jacobian**: the largest gap between the stated log Jacobian determinant
  and a finite-difference one;
- **onto**: the largest error in mapping random interior tables to parameters
  and back;
- **unreachable**: the share of random tables with no parameter vector that
  produces them.

The table gives the worst case for each scheme over all sizes and orders.

| Scheme | Margins | Smallest cell | Inverse | Jacobian | Onto | Unreachable |
|---|---|---|---|---|---|---|
| between bounds, sharp | 3.7e-16 | 6.6e-07 | 3.7e-12 | 1.0e-08 | 1.9e-16 | 0.0% |
| between bounds, smoothed (delta 0.05) | 3.4e-16 | 8.0e-07 | 5.0e-12 | 4.9e-08 | 3.4e-16 | 1.0% |
| cell logit | 3.4e-16 | 3.4e-06 | 1.2e-12 | 1.1e-08 | 6.9e-15 | 0.0% |
| adjusted row | 3.7e-16 | 6.4e-08 | 2.1e-12 | 2.4e-08 | 2.6e-16 | 0.0% |
| adjusted table | 5.7e-16 | 1.3e-06 | 1.8e-15 | 1.7e-08 | 1.0e-15 | 0.0% |


Every scheme reproduces the margins to rounding error, returns strictly
positive tables, inverts exactly and has a correct Jacobian. The one
exception is the smoothed between-bounds scheme, where about 1% of random
tables are unreachable because smoothing the kinks pulls the cell away
from the Fréchet bounds. The sharp version has no such gap.


## Extensions to arrays

@Chen2006 and @Dinwoodie2011 describe sequential importance sampling methods for more complex cases, in particular sampling from multi-way tables. I have not yet tried to implement these in Stan but imagine it should be possible to do so.

## An example: votes for parties and candidates

```{r define_tbl}
#| echo: false
#| warning: false
#| message: false
s60_rm <- c("con" = 1721, "lab" = 12111, "lib" = 1570, "other" = 525, "green" = 281, "SNP" = 6998, "spoil" = 667)
s60_cm <- c("con" = 1648, "lab" = 9437, "lib" = 1494, "other" = 2708, "green" = 580, "SNP" = 7198, "spoil" = 808)
actual_cell_60 <- structure(list(row_no = c(1, 2, 3, 6, 7, 4, 5, 1, 2, 3, 6, 7,
                          4, 5, 1, 2, 3, 6, 7, 4, 5, 1, 2, 3, 6, 7, 4, 5, 1, 2, 3, 6, 7,
                          4, 5, 1, 2, 3, 6, 7, 4, 5, 1, 2, 3, 6, 7, 4, 5),
               col_no = c(1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 4, 5, 5, 5, 5, 5, 5, 5, 6, 6, 6, 6, 6, 6, 6, 7, 7, 7, 7, 7, 7, 7),
               actual_cell_value = c(1202, 277, 62, 54, 36, 6, 11, 29, 8941, 61, 129, 203, 53, 21, 25, 320, 1023, 74, 28, 19, 5, 353, 1076, 183, 502, 97, 328, 169, 27, 225, 121, 122, 10, 38, 37, 46, 771, 105, 6007, 168, 71, 30, 39, 501, 15, 110, 125, 10, 8)), row.names = c(NA, -49L), class = c("tbl_df", "tbl", "data.frame"))
```

```{r formattbl}
name_lookup <- function(x){
  case_when(
    x == 1 ~ "con",
    x == 2 ~ "lab",
    x == 3 ~ "lib",
    x == 4 ~ "other",
    x == 5 ~ "green",
    x == 6 ~ "SNP",
    x == 7 ~ "spoil"
  )
}

n_diag <- min(length(unique(actual_cell_60$row_no)), 
              length(unique(actual_cell_60$col_no)))
diag_coords <- cbind(1:n_diag + 1, 1:n_diag + 1)

```

```{r margin_tbl}
#| tbl-cap: "Marginal Total Votes in the 2008 Scottish Parliament Election for party lists (rows) and candidates with party labels (columns). Substantively we know that a large volume of the vote splits in this table will sit on the diagonal with voters choosing the same party with both votes."

ht_blank <- actual_cell_60 |>
  mutate(
    row = name_lookup(row_no),
    col = name_lookup(col_no)
  ) |>
  arrange(col_no, row_no) |>
  select(actual_cell_value, row, col) |> 
  pivot_wider(names_from = col, values_from = actual_cell_value) |>
  adorn_totals(where = c("row", "col")) |>
  # rename(" " = row) |>
  mutate(across(-c(row, Total), \(x) ifelse(row == "Total", x, ""))) |>
  as_huxtable() |>
  set_font_size(8) |>
  set_bold(1, everywhere, TRUE) |>
  set_bold(everywhere, 1, TRUE) |>
  set_bottom_border(1, everywhere, 0.4) |>
  set_bottom_border(final(1), everywhere, 0.4) |>
  set_bold(final(1), everywhere, TRUE) |>
  set_bold(everywhere, final(1), TRUE)|>
  set_all_padding(3)

background_color(ht_blank)[diag_coords] <- "#FFF2AC"
number_format(ht_blank) <- list(function(x) format(x, big.mark = ",", scientific = FALSE)) 

ht_blank
```


```{r full_tbl}
#| echo: false
#| warning: false
#| message: false
#| tbl-cap: "Votes in the 2008 Scottish Parliament Election for party lists (rows) and candidates with party labels (columns)."

ht <- actual_cell_60 |>
  mutate(row = name_lookup(row_no),
         col = name_lookup(col_no)) |>
  arrange(col_no, row_no) |>
  select(actual_cell_value, row, col) |> 
  pivot_wider(names_from = col, values_from = actual_cell_value) |>
  adorn_totals(where = c("row", "col")) |>
  rename(" "=row)|>
  as_huxtable() |>
  set_bold(1, everywhere, TRUE) |>              # Bold column headers
  set_bottom_border(1, everywhere, 0.4) |>      # Top rule below header
  set_bottom_border(final(1), everywhere, 0.4) |># Bottom rule above totals
  set_bold(final(1), everywhere, TRUE) |>        # Bold total row
  set_bold(everywhere, final(1), TRUE)  |>        # Bold total column
  set_bold(everywhere, 1, TRUE) |> 
  set_font_size(9) |>
  set_all_padding(3)
background_color(ht)[diag_coords] <- "#FFF2AC"
number_format(ht) <- list(function(x) format(x, big.mark = ",", scientific = FALSE)) 

ht

```
